Storm surge modelling in the Strait of Georgia

### 2014-12-11 RoughWork/manuscript.tex
a few typoographical edits before sending to east coast

This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its accuracy in simulating storm surge hindcasts. This baroclinic model includes the effect of stratification and rivers, however, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep and so wetting and drying effects are believed to be unimportant. A detailed overview of the model configuration and domain including a description of the ocean model configuration is provided in section \ref{sec:config}. This section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hindcasts is also assessed in section \ref{sec:storm}. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

Temperature and salinity at the Juan de Fuca boundary were taken from a weekly climatology which was created from results from a model covering the Salish Sea and the west coast of Vancouver Island \citep{massonfine2012}.  Their results, originally on $s$-levels were interpolated onto $z$-levels and then onto the NEMO horizontal grid.  To prepare the climatology all years (1995-2008) were averaged and results, approximately every 15 days, were interpolated to a weekly climatology. The model is relaxed to the forced temperature and salinity over the 10 grid points (about 5km) closest to the open boundaries, using a flow relaxation scheme \citep{engedahl1995use}. %other references?

Sea surface height at the mouth of Juan de Fuca was set using values from the Tofino tide gauge \citep{DFOObservations} which is outside of, but close to, the western boundary of the model domain.  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values, de-tided and with the zero reset as for the climatology. The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait. The tidal forcing and sea surface height was applied to the barotropic velocity forcing with the NEMO Flather scheme \citep{flather1994storm, madec2012nemo}. At the west, the baroclinic velocities at the boundary were set equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO 3.4, the version used here.  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain. At the north baroclinic velocities were set to zero.

@@

River input provides a significant volume of freshwater to the Salish Sea and can influence stratification, circulation, and primary productivity. However, most rivers in the domain are not gauged so parametrizations were required to represent river flow. \citet{morrison2011rivers} provides a method for estimating freshwater runoff in the Salish Sea region based on precipitation. Monthly runoff volumes for each watershed for each year from 1970 to 2012 were acquired from \citet{morrison2011rivers}, as well as monthly averages.

@@

Modelled residuals are generally too low in this hindcast (Figure \ref{fig:dec2006}) however the timing of the surge agrees well. All Willmott skill scores rank above 0.88 and the $\gamma^2$ measure also indicates good agreement between model and observations (Table \ref{tab:statistics}) but the performance has degraded when compared to the Feb 2006 and Nov 2006 cases. This could be due to a number of factors. First, the very large wind speeds observed during this storm are significantly underrepresented by the atmospheric model leading to lower wind stress and less piling of water against the coast. Second, the anomaly forcing at the open boundaries could be too low. The observed residual at Tofino, BC (a point near by but outside of the domain) is up to 10 cm lower than the observed residual at Neah Bay, WA (a point near by but inside of the domain). Since the storm surge is less sensitive to wind forcing at the surface than anomaly forcing at the open boundaries, the latter is likely the cause.

@@

Hindcasts with both atmospheric forcing products agree well with the observations (Figure \ref{fig:dec2012_surge}). Willmott skill scores for both modelled residuals are above 0.92 at these four locations. Further, the modelled residuals deviate only slightly from each other, most noticeable at Point Atkinson during the large wind event on Dec 15, 2012 where simulation with the low resolution atmospheric forcing has a residual that is 5 cm higher. Although the wind speeds on Dec 15, 2012 are quite high, this event did not result in any flooding or residuals greater than 50 cm. Significant flooding occurred later, on Dec 17, 2012, where the observed surge amplitude at Point Atkinson was 68.3 cm. Even though the low resolution model significantly under predicts the wind speed at some locations, it performs comparable to forcing with high resolution products. This result further demonstrates that the local wind is not the most important factor for setting up storm surges over this domain.

>>>>

This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its accuracy in simulating storm surge hindcasts. This baroclinic model includes the effect of stratification and rivers, however, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep and so wetting and drying effects are believed to be unimportant. A detailed overview of the model configuration and domain is provided in section \ref{sec:config}. This section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hindcasts is also assessed in section \ref{sec:storm}. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

Temperature and salinity at the Juan de Fuca boundary were taken from a weekly climatology which was created from results from a model covering the Salish Sea and the west coast of Vancouver Island \citep{massonfine2012}.  Their results, originally on $s$-levels were interpolated onto $z$-levels and then onto the NEMO horizontal grid.  To prepare the climatology all years (1995-2008) were averaged and results, approximately every 15 days, were interpolated to a weekly climatology. The model is relaxed to the forced temperature and salinity over the 10 grid points (about 5 km) closest to the open boundaries, using a flow relaxation scheme \citep{engedahl1995use}. %other references?

Sea surface height at the mouth of Juan de Fuca was set using values from the Tofino tide gauge \citep{DFOObservations} which is outside of, but close to, the western boundary of the model domain.  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values with the tides removed. Tidal predictions were generated with a tidal analysis package called t\_tide \citep{pawlowicz2002classical} using a time series from the year prior to the surge event to avoid over predicted constituents during high surge years. The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait. The tidal forcing and sea surface height was applied to the barotropic velocity forcing with the NEMO Flather scheme \citep{flather1994storm, madec2012nemo}. At the west, the baroclinic velocities at the boundary were set equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO 3.4, the version used here.  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain. At the north baroclinic velocities were set to zero.

@@

River input provides a significant volume of fresh water to the Salish Sea and can influence stratification, circulation, and primary productivity. However, most rivers in the domain are not gauged so parametrizations were required to represent river flow. \citet{morrison2011rivers} provides a method for estimating freshwater runoff in the Salish Sea region based on precipitation. Monthly runoff volumes for each watershed for each year from 1970 to 2012 were acquired from \citet{morrison2011rivers}, as well as monthly averages.

@@

Modelled residuals are generally too low in this hindcast (Figure \ref{fig:dec2006}), however, the timing of the surge agrees well. All Willmott skill scores rank above 0.88 and the $\gamma^2$ measure also indicates good agreement between model and observations (Table \ref{tab:statistics}) but the performance has degraded when compared to the Feb 2006 and Nov 2006 cases. This could be due to a number of factors. First, the very large wind speeds observed during this storm are significantly underrepresented by the atmospheric model leading to lower wind stress and less piling of water against the coast. Second, the anomaly forcing at the open boundaries could be too low. The observed residual at Tofino, BC (a point near by but outside of the domain) is up to 10 cm lower than the observed residual at Neah Bay, WA (a point near by but inside of the domain). Since the storm surge is less sensitive to wind forcing at the surface than anomaly forcing at the open boundaries, the latter is likely the cause.

@@

Hindcasts with both atmospheric forcing products agree well with the observations (Figure \ref{fig:dec2012_surge}). Willmott skill scores for both modelled residuals are above 0.92 at these four locations. Further, the modelled residuals deviate only slightly from each other, most noticeable at Point Atkinson during the large wind event on Dec 15, 2012 where simulation with the low resolution atmospheric forcing has a residual that is 5 cm higher. Although the wind speeds on Dec 15, 2012 are quite high, this event did not result in any flooding or residuals greater than 50 cm. Significant flooding occurred later, on Dec 17, 2012, where the observed surge amplitude at Point Atkinson was 68.3 cm. Even though the low resolution model significantly under predicts the wind speed at some locations, it performs comparably to forcing with high resolution products. This result further demonstrates that the local wind is not the most important factor for setting up storm surges over this domain.
